---
title: "p8105_hw2_rl3401.Rmd"
author: "Ruoxi Li"
date: "2023-09-26"
output: github_document
---
```{r,echo=FALSE}
Sys.setenv(LANGUAGE = "en")
```

```{r}
library(tidyverse)
library(readxl)
```


## Problem 1

Clean the data in pols-month.csv

```{r}
pols_df = read_csv("data/pols-month.csv") |>
  janitor::clean_names() |>
  separate(mon, into = c("year", "month", "day"), sep = "-") |>
  mutate(month = case_when(
    month == "01" ~ "jan",
    month == "02" ~ "feb",
    month == "03" ~ "mar",
    month == "04" ~ "apr",
    month == "05" ~ "may",
    month == "06" ~ "jun",
    month == "07" ~ "jul",
    month == "08" ~ "aug",
    month == "09" ~ "sep",
    month == "10" ~ "oct",
    month == "11" ~ "nov",
    month == "12" ~ "dec",
  )) |>
  mutate(president = case_match(
    prez_dem,
    0 ~ "gop",
    1 ~ "dem"
  )) |>
  select(-prez_dem, -prez_gop) |>
  select(-day)
```

Clean the data in snp.csv 

```{r}
date_format <- "%m/%d/%y"
snp_df = read_csv("data/snp.csv", col_types = cols(date = col_date(format = date_format))) |>
  janitor::clean_names() |>
  separate(date, into = c("year", "month", "day"), sep = "-") |>
   mutate(month = case_when(
    month == "01" ~ "jan",
    month == "02" ~ "feb",
    month == "03" ~ "mar",
    month == "04" ~ "apr",
    month == "05" ~ "may",
    month == "06" ~ "jun",
    month == "07" ~ "jul",
    month == "08" ~ "aug",
    month == "09" ~ "sep",
    month == "10" ~ "oct",
    month == "11" ~ "nov",
    month == "12" ~ "dec",
  )) |>
    select(-day) |>
  arrange(year,month) |>
  relocate(year,month)  
```

Tidy the unemployment data 

```{r}
unemployment_df = read_csv("data/unemployment.csv") |>
   janitor::clean_names() |>
   mutate(year = as.character(year))|>
   pivot_longer(
    jan:dec,
    names_to = "month",
    values_to = "unemployment_rate",
  ) 
```

Join the datasets by merging snp into pols, and merging unemployment into the result

```{r}
final_df = 
  left_join(pols_df, snp_df, by = c("year","month"))|>
  left_join(unemployment_df, by = c("year","month") )
```

*The pols_df has `r nrow(pols_df)` rows and `r ncol(pols_df)` columns.*

*The snp_df has `r nrow(snp_df)` rows and `r ncol(snp_df)` columns.*

*The unemployment_df has `r nrow(unemployment_df)` rows and `r ncol(unemployment_df)` columns.*

*The resulting dataset has has `r nrow(final_df)` rows and `r ncol(final_df)` columns with variables `r names(final_df)`.*

## problem 2

Read and clean the Mr. Trash Wheel sheet

```{r}
mr_trash_df = 
read_excel("data/TWC.xlsx", sheet = "Mr. Trash Wheel", skip = 1, range = cell_cols("A:N")) |>
  janitor::clean_names()  |>
  drop_na(dumpster) |>
  mutate(
    year=as.integer(year),
    homes_powered = weight_tons*500/30,
    wheel_type = "mr")
```

Import, clean, and organize the data for Professor Trash Wheel and Gwynnda


```{r}
prof_trash_df = 
  read_excel("data/TWC.xlsx", sheet = "Professor Trash Wheel", skip = 1, range = cell_cols("A:M")) |>
  janitor::clean_names()  |>
  drop_na(dumpster) |>
  mutate(
    year=as.integer(year),
    homes_powered = weight_tons*500/30,
    wheel_type = "prof")

gwy_trash_df = 
  read_excel("data/TWC.xlsx", sheet = "Gwynnda Trash Wheel", skip = 1, range = cell_cols("A:K")) |>
  janitor::clean_names()  |>
  drop_na(dumpster) |>
  mutate(
    year=as.integer(year),
    homes_powered = weight_tons*500/30,
    wheel_type = "gwy")
```

Combine datasets to produce a single tidy dataset

```{r}
trash_df = 
  bind_rows(mr_trash_df, prof_trash_df, gwy_trash_df) |>
  relocate(wheel_type)
```

*The Mr. Trash Wheel dataframe has `r nrow(mr_trash_df)` rows and `r ncol(mr_trash_df)` columns.*

*The Professor dataframe Trash Wheel has `r nrow(prof_trash_df)` rows and `r ncol(prof_trash_df)` columns.*

*The Gwynnda dataframe has `r nrow(gwy_trash_df)` rows and `r ncol(gwy_trash_df)` columns.*

*The combined dataframe has `r nrow(trash_df)` rows and `r ncol(trash_df)` columns.*

*The example of key variables in the combined dataframe is*

```{r,echo=FALSE,message=FALSE,warning=FALSE}
head(trash_df,n=1)
```
*The total weight of trash collected by Professor trash wheel is `r sum(pull(prof_trash_df,weight_tons))` tons.*

*The total number of cigarette butts collected by Gwynnda in July of 2021 is `r gwy_trash_df|> filter(year == 2021, month == "July") |> pull(cigarette_butts) |> sum() `.*

## problem 3

Import, clean, and tidy the dataset of baseline demographics

```{r}
baseline_df = 
read_csv("data/MCI_baseline.csv", skip = 1,show_col_types = FALSE) |>
  janitor::clean_names()  |>
  filter(age_at_onset == "." | age_at_onset > current_age)|>
   mutate(
      sex = case_match(
      sex,
      1 ~ "male",
      0 ~ "female"),
      apoe4  = case_match(
      apoe4,
      1 ~ "carrier",
      0 ~ "non_carrier"),
      current_age = as.numeric(current_age)
  ) 
```

*There are `r nrow(baseline_df)` participants recruited, `r nrow(baseline_df)-nrow(filter(baseline_df,age_at_onset == "."))` develop MCI.*

*The average baseline age is `r pull(baseline_df,current_age) |> mean()`.*

*The proportion of women in the study are APOE4 carriers is `r  nrow(filter(baseline_df,sex == "female", apoe4 == "carrier"))/nrow(filter(baseline_df,sex == "female"))`.*

```{r}
amyloid_df = 
read_csv("data/mci_amyloid.csv", skip = 1,show_col_types = FALSE) |>
  janitor::clean_names()  |>
  rename("id" = "study_id")
```

*I read the data from a csv file, omitted the row with comment, and clean the columne names. I renamed the study_id column to id column to further combine the datasets.*

Check whether some participants appear in only the baseline or amyloid dataset

```{r}
full_ad_df = full_join(baseline_df,amyloid_df,by = "id")
```

*After checking the dataframe, I found that some participants appear in only the baseline dataset, while some other participants appear in only the amyloid dataset.*

Combine the demographic and biomarker datasets so that only participants who appear in both datasets are retained

```{r}
inner_ad_df = inner_join(baseline_df,amyloid_df,by = "id")
```

*Only `r nrow(inner_ad_df)` participants appear in both the baseline dataset and the amyloid dataset, meaning that there are `r nrow(full_ad_df)-nrow(inner_ad_df)` participants whose data only appear in one dataset.*
 
*The resulting dataset of participants who appear in both baseline and amyloid datasets have `r nrow(inner_ad_df)` rows and `r ncol(inner_ad_df)` columns, the variables in the dataset are: `r names(inner_ad_df)`.*


Export the result as a CSV to the data directory

```{r}
write_csv(inner_ad_df, "data/inner_ad_df.csv")
```


