---
title: "p8105_hw2_rl3401.Rmd"
author: "Ruoxi Li"
date: "2023-09-26"
output: github_document
---
```{r,echo=FALSE}
Sys.setenv(LANGUAGE = "en")
```

```{r}
library(tidyverse)
library(readxl)
```


## Problem 1

clean the data in pols-month.csv

```{r}
pols_df = read_csv("data/pols-month.csv") |>
  janitor::clean_names() |>
  separate(mon, into = c("year", "month", "day"), sep = "-") |>
  mutate(month = case_when(
    month == "01" ~ "jan",
    month == "02" ~ "feb",
    month == "03" ~ "mar",
    month == "04" ~ "apr",
    month == "05" ~ "may",
    month == "06" ~ "jun",
    month == "07" ~ "jul",
    month == "08" ~ "aug",
    month == "09" ~ "sep",
    month == "10" ~ "oct",
    month == "11" ~ "nov",
    month == "12" ~ "dec",
  )) |>
  mutate(president = case_match(
    prez_dem,
    0 ~ "gop",
    1 ~ "dem"
  )) |>
  select(-prez_dem, -prez_gop) |>
  select(-day)
```

clean the data in snp.csv 

```{r}
date_format <- "%m/%d/%y"
snp_df = read_csv("data/snp.csv", col_types = cols(date = col_date(format = date_format))) |>
  janitor::clean_names() |>
  separate(date, into = c("year", "month", "day"), sep = "-") |>
   mutate(month = case_when(
    month == "01" ~ "jan",
    month == "02" ~ "feb",
    month == "03" ~ "mar",
    month == "04" ~ "apr",
    month == "05" ~ "may",
    month == "06" ~ "jun",
    month == "07" ~ "jul",
    month == "08" ~ "aug",
    month == "09" ~ "sep",
    month == "10" ~ "oct",
    month == "11" ~ "nov",
    month == "12" ~ "dec",
  )) |>
    select(-day) |>
  arrange(year,month) |>
  relocate(year,month)  
```

tidy the unemployment data 

```{r}
unemployment_df = read_csv("data/unemployment.csv") |>
   janitor::clean_names() |>
   mutate(year = as.character(year))|>
   pivot_longer(
    jan:dec,
    names_to = "month",
    values_to = "unemployment_rate",
  ) 
```

Join the datasets by merging snp into pols, and merging unemployment into the result

```{r}
final_df = 
  left_join(pols_df, snp_df, by = c("year","month"))|>
  left_join(unemployment_df, by = c("year","month") )
```

Write a short paragraph about these datasets. Explain briefly what each dataset contained, and describe the resulting dataset (e.g. give the dimension, range of years, and names of key variables).

## problem 2

Read and clean the Mr. Trash Wheel sheet

```{r}
mr_trash_df = 
read_excel("data/TWC.xlsx", sheet = "Mr. Trash Wheel", skip = 1, range = cell_cols("A:N")) |>
  janitor::clean_names()  |>
  drop_na(dumpster) |>
  mutate(
    year=as.integer(year),
    homes_powered = weight_tons*500/30,
    wheel_type = "mr")
```

import, clean, and organize the data for Professor Trash Wheel and Gwynnda


```{r}
pro_trash_df = 
  read_excel("data/TWC.xlsx", sheet = "Professor Trash Wheel", skip = 1, range = cell_cols("A:M")) |>
  janitor::clean_names()  |>
  drop_na(dumpster) |>
  mutate(
    year=as.integer(year),
    homes_powered = weight_tons*500/30,
    wheel_type = "prof")

gwy_trash_df = 
  read_excel("data/TWC.xlsx", sheet = "Gwynnda Trash Wheel", skip = 1, range = cell_cols("A:K")) |>
  janitor::clean_names()  |>
  drop_na(dumpster) |>
  mutate(
    year=as.integer(year),
    homes_powered = weight_tons*500/30,
    wheel_type = "gwy")
```

Combine datasets to produce a single tidy dataset

```{r}
trash_df = 
  bind_rows(mr_trash_df, pro_trash_df, gwy_trash_df) |>
  relocate(wheel_type)
```

Write a paragraph about these data; you are encouraged to use inline R. Be sure to note the number of observations in the resulting dataset, and give examples of key variables. For available data, what was the total weight of trash collected by Professor Trash Wheel? What was the total number of cigarette butts collected by Gwynnda in July of 2021?

## problem 3

Import, clean, and tidy the dataset of baseline demographics

```{r}
baseline_df = 
read_csv("data/MCI_baseline.csv", skip = 1,show_col_types = FALSE) |>
  janitor::clean_names()  |>
  filter(age_at_onset == "." | age_at_onset > current_age)|>
   mutate(
      sex = case_match(
      sex,
      1 ~ "male",
      0 ~ "female"),
      apoe4  = case_match(
      apoe4,
      1 ~ "carrier",
      0 ~ "non_carrier"),
      current_age = as.numeric(current_age)
  ) 
```

Discuss important steps in the import process and relevant features of the dataset. How many participants were recruited, and of these how many develop MCI? What is the average baseline age? What proportion of women in the study are APOE4 carriers?

import, clean, and tidy the dataset of longitudinally observed biomarker values

```{r}
amyloid_df = 
read_csv("data/mci_amyloid.csv", skip = 1,show_col_types = FALSE) |>
  janitor::clean_names()  |>
  rename("id" = "study_id")
```

comment on the steps on the import process and the features of the dataset.

Check whether some participants appear in only the baseline or amyloid dataset

```{r}
full_ad_df = full_join(baseline_df,amyloid_df,by = "id")
```

After checking the dataframe, I found that some participants appear in only the baseline dataset, while some other participants appear in only the amyloid dataset.
Check whether some participants appear in only the baseline or amyloid datasets, and comment on your findings. 

Combine the demographic and biomarker datasets so that only participants who appear in both datasets are retained

```{r}
inner_ad_df = inner_join(baseline_df,amyloid_df,by = "id")
```

describe the resulting dataset

export the result as a CSV to the data directory

```{r}
write_csv(inner_ad_df, "data/inner_ad_df.csv")
```


